Explore the SingleR results

### 2025-07-16 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
add final plot and helper function for subsetting nbatlas marker genes

# Source additional utilities functions:
# - harmonize_celltypes()
# - faceted_umap()
# - generate_dotplot()
source(file.path(module_dir, "scripts", "utils", "celltype-utils.R"))
```

@@

label_map_df |>
  dplyr::group_by(label_recoded) |>
  dplyr::summarize(all_labels = paste(label, collapse = ", ")) |>
  dplyr::arrange(label_recoded) |>
  dplyr::rename(
    "Label shown in figures" = label_recoded,
    "Underlying NBAltas label(s)" = all_labels
  ) |>
  knitr::kable()
```

@@

```{r, fig.height = 10, fig.width = 10} 
# Pivot wider for heatmap functions
celltype_recoded_wide_df <- celltype_recoded_df |>
  tidyr::pivot_wider(
    names_from = annotation_type,
    values_from = label_recoded
  )

@@

```{r}
# number of marker genes to consider per validation group
n_marker_genes <- 5

top_nbatlas_markers_df <- nbatlas_markers_df |>
  # keep only the top `n_marker_genes`
  dplyr::filter(direction == "up") |>
  dplyr::group_by(NBAtlas_label) |>
  dplyr::mutate(rank_lfc = rank(-avg_log2FC)) |>
  dplyr::ungroup() |>
  dplyr::filter(rank_lfc <= n_marker_genes) |>
  dplyr::select(marker_gene_label = NBAtlas_label, gene_symbol, ensembl_gene_id)
```

@@

Next, we look at expression of consensus validation marker genes for each normal/harmonized cell type label.
We compare expression across groups of cells:

* `matching annotations`: Cells whose consensus annotation matches the `SingleR` annotation
* `different annotations`: Cells whose consensus annotation differs from the `SingleR` annotation
* `consensus not classified`: Cells where the consensus annotation is unknown

Each panel contains only cells that `SingleR` annotated with the given label, and shows only marker genes associated with that given cell type. 
We expect that cells whose consensus and `SingleR` annotations agree may show higher gene expression for the given marker.

```{r}
# this time we are not considering putative tumor
direct_celltype_matches <- c(
  "B cell",
  "T cell",
  # note that we need to collapse the consensus to compare
  "myeloid",
  "fibroblast",
  "endothelial cell",
  "plasma cell",
  "natural killer cell", 
  "unknown_consensus" # we do want to keep these for this analysis
)

# Define the different myeloid types in consensus annotations
consensus_myeloid <- c("myeloid", "dendritic cell", "monocyte", "macrophage")

# Data frame with column `compare_annotations` that indicates
# if consensus and singler (roughly) match
compare_annotations_df <- celltype_recoded_df |>
  # recode labels for comparison
  dplyr::mutate(
    label_recoded = ifelse(
      label_recoded %in% consensus_myeloid,
      "myeloid",
      label_recoded
  )) |>
  dplyr::filter(label_recoded %in% direct_celltype_matches) |>
  dplyr::select(cell_id, annotation_type, label_recoded) |>
  # make it wider to label the situation
  tidyr::pivot_wider(
    names_from = "annotation_type",
    values_from = "label_recoded"
  ) |>
  # remove NAs:
  # - consensus NAs are those with other cell types not to include here
  # - singler NAs were unclassified cells
  tidyr::drop_na() |>
  dplyr::mutate(
    compare_annotations = dplyr::case_when(
      consensus_validation == "unknown_consensus" ~ "consensus not classified", 
      consensus_validation == singler ~ "matching annotations",
      consensus_validation != singler ~ "different annotations"
    ), 
    compare_annotations = forcats::fct_relevel(compare_annotations, "matching annotations", "different annotations", "consensus not classified")
  ) |>
  # we can remove the consensus annotation now
  dplyr::select(cell_id, singler, compare_annotations)

direct_markers_df <- consensus_markers_df |>
  dplyr::filter(validation_group_annotation %in% direct_celltype_matches) |>
  dplyr::select(marker_gene_label = validation_group_annotation, ensembl_gene_id, gene_symbol)
cells <- unique(compare_annotations_df$cell_id)
genes <- unique(direct_markers_df$ensembl_gene_id)
```

```{r}
compare_annotations_expr_df <- scuttle::makePerCellDF(
  merged_sce[genes, cells],
  features = genes,
  assay.type = "logcounts",
  use.coldata = "cell_id",
  use.dimred = FALSE
) |>
  tidyr::pivot_longer(starts_with("ENSG"), names_to = "ensembl_gene_id", values_to = "logcounts") |>
  # filter to only cells with expression
  dplyr::filter(logcounts > 0) |>
  # join other data frames
  dplyr::left_join(compare_annotations_df, by = "cell_id") |>
  dplyr::left_join(direct_markers_df, by = "ensembl_gene_id", relationship = "many-to-many")
```

```{r, fig.width = 10, fig.height = 20}
unique(compare_annotations_expr_df$singler) |>
  purrr::map(
    \(celltype) {

      plot_df <- compare_annotations_expr_df |>
        dplyr::filter(singler == celltype, marker_gene_label == celltype)

      ggplot(plot_df) + 
        aes(x = gene_symbol, y = logcounts, fill = compare_annotations) +
        geom_boxplot(outlier.size = 0.25) + 
        ggtitle(glue::glue("cells annotated as: {celltype}"))

  }) |>
  patchwork::wrap_plots(ncol = 1) + patchwork::plot_layout(guides = "collect")
```

In most cases, gene expression for cells with matching consensus/`SingleR` annotations is higher or about the same as is gene expression for cells with different annotations.

>>>>

# Source additional utilities functions:
# - select_nbatlas_markers()
# - harmonize_celltypes()
# - faceted_umap()
# - generate_dotplot()
source(file.path(module_dir, "scripts", "utils", "celltype-utils.R"))
```

@@

label_map_df |>
  dplyr::group_by(label_recoded) |>
  dplyr::summarize(all_labels = paste(label, collapse = ", ")) |>
  dplyr::arrange(label_recoded) |>
  dplyr::rename(
    "Label shown in figures" = label_recoded,
    "Underlying NBAltas label(s)" = all_labels
  )
```

@@

```{r, fig.height = 8, fig.width = 8} 
# Pivot wider for heatmap functions
celltype_recoded_wide_df <- celltype_recoded_df |>
  dplyr::select(-label) |>
  tidyr::pivot_wider(
    names_from = annotation_type,
    values_from = label_recoded
  )

@@

```{r}
# data frame of top 5 upregulated genes per cell type
top_nbatlas_markers_df <- subset_nbatlas_markers(nbatlas_markers_df, 5, "up")
```

@@

Next, we look at expression of consensus validation marker genes for each normal/harmonized cell type label.
We're specifically interested in the marker gene expression for cells that were classified by `SingleR` but have an unknown consensus cell type.
We assume the unknown consensus cells may be more likely to be tumor because they could not be robustly classified otherwise, but since some of those cells were called "normal" by `SingleR` we'd like to get a sense of whether they really are normal.

We have two groups:

* `SingleR` cells classified as `Neuroendocrine`; these are putatively tumor cells
  * We expect that these will have higher `Neuroendocrine` marker gene expression and lower normal gene expression
* `SingleR` cells not classified as `Neuroendocrine`; these are more likely to be normal cells
  * We expect that these will have higher normal marker gene expression

For this, we'll consider the average gene expression among the top 20 NBAtlas marker genes for each cell type.

```{r}
# data frame of top 20 upregulated genes per cell type
top_nbatlas_markers_df <- subset_nbatlas_markers(nbatlas_markers_df, 20, "up")

unknown_consensus_ids <- celltype_recoded_wide_df |>
  dplyr::filter(consensus_validation == "unknown_consensus") |>
  dplyr::pull(cell_id)

unknown_consensus_celltype_df <- full_celltype_df |>
  dplyr::filter(cell_id %in% unknown_consensus_ids) |>
  dplyr::mutate(
    singler = ifelse(
      label == "Neuroendocrine", "NE (likely tumor)", "Non-NE (likely normal)")
  ) |>
  dplyr::select(-label)

goi <- unique(top_nbatlas_markers_df$ensembl_gene_id)

compare_annotations_expr_df <- scuttle::makePerCellDF(
  merged_sce[goi, unknown_consensus_ids],
  features = goi,
  assay.type = "logcounts",
  use.coldata = "cell_id",
  use.dimred = FALSE
) |>
  tidyr::pivot_longer(starts_with("ENSG"), names_to = "ensembl_gene_id", values_to = "logcounts") |>
  # filter to only cells with expression
  dplyr::filter(logcounts > 0) |>
  # join the annotations and marker gene data frame; use inner since we've filtered already
  dplyr::inner_join(unknown_consensus_celltype_df, by = "cell_id") |>
  dplyr::left_join(top_nbatlas_markers_df, by = "ensembl_gene_id", relationship = "many-to-many") |>
  # average by marker_gene_label
  dplyr::group_by(singler, marker_gene_label, ensembl_gene_id) |>
  dplyr::summarize(mean_exp = mean(logcounts)) |>
  dplyr::ungroup() |>
  dplyr::mutate(marker_gene_label = factor(marker_gene_label), 
                marker_gene_label = forcats::fct_relevel(marker_gene_label, nbatlas_bar_order))

```{r, fig.width = 12, fig.height = 5}
ggplot(compare_annotations_expr_df) + 
  aes(x = marker_gene_label, y = mean_exp, color = singler) + 
  ggforce::geom_sina(position = position_dodge(width = 0.7)) + 
  labs(
    x = "NBAtlas marker gene label", 
    y = "Mean expression across marker genes",
    color = "SingleR annotation"
  ) + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
The left-most gene set on the x-axis is `Neuroendocrine`, and normal cell types follow.
This plot suggests that the normal cells `SingleR` is picking up are likely to be normal cells: normal gene expression for non-NE cells is higher compared to NE cells, and we see the opposite trend for the `Neuroendocrine` marker genes.
